---
title: "wlsd Version History"
author: Charles Ingulli
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette
nocite: '@*'
vignette: >
  %\VignetteIndexEntry{wlsd}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = FALSE,
  comment = "#>"
)
```

## Overview

A table guide to summarize changes between versions.

```{r, echo = FALSE, results = "asis"}
VH <- data.frame(Version = c("0.0.1.1","0.0.1.2","0.0.2"),
                 Date = c("9/8/2020","10/16/2020","12/1/2020"),
                 Author = c("CI","CI","CI"),
                 Summary = c("Initial version of package.","Update data and functions.","Update data set.")
                 )
knitr::kable(VH, caption = "Table of Changes for wlsd Versions")
```


## Development Version 0.0.2
1. Update LBP data set so variables are parsed in proper formats.

## (Development) Version 0.0.1.2 

1. Update the data set.
2. Add `takefirst()` function.


## (Development) Version 0.0.1.1 

1. Add the base version of the package with important functions.
2. Add basic package usage vignette. 
3. Add first version history vignette.
